---
title: "Application Activity"
subtitle: "Unit 1 Review"
author: "Nathan Lunceford"
format:
  html:
    self-contained: true
    code-fold: true
---

# Assessment Objectives

  * Apply the 3 rules of probability in different scenarios
  * Describe the 5 steps of the statistical process
  * Differentiate between an observational study and an experiment
  * Differentiate between a population and a sample
  * Describe the 4 random sampling schemes
  * Explain importance of random sampling
<br/>
  * Distinguish between quantitative and categorical variables
  * Create graphical summaries of data (boxplots, histograms) and be able to make conclusions based on the graphs
  * Create numerical summaries of data (mean, sd, 5 number summary, percentiles) and compare groups (centers, spreads)
  * Distinguish between a parameter and a statistic
  * Determine the 5 number summary from a boxplot
<br/>
  * State the properties of a normal distribution
  * Calculate the z-score for an individual from a normal population with mean, $\mu$, and std. dev, $\sigma$
  * Calculate probabilities (>, <, between) of certain values given $\mu$ and $\sigma$
<br/>
  * Describe the concept of a sampling distribution of the sample mean
  * State the Central Limit Theorem
  * Determine the mean of the sampling distribution of the sample mean given $\mu$ and $\sigma$
  * Determine the std. dev. of the sampling distribution of the sample mean given $\mu$ and $\sigma$
  * Determine the _shape_ of the sampling distribution given a sample size, n, and $\mu$ and $\sigma$
  * State the Law of Large Numbers
  
  
# Design the Study

You and a colleague are interested in studying the relationship between social media use and mental health among youth ages 18-24. You decide to use the Psychological Wellbeing Scale (PWB) developed by Stanford University as your metric for comparison (the response variable).

__Q1.1:__ 

List 5 factors (explanatory variables) related to social media use that might impact Wellbeing.  

Include whether or not the factors are quantitative or categorical (include at least one of each):

1. **Time spent on social media (Quantitative):** The total number of hours per day spent on social media platforms.
2. **Type of social media platforms used (Categorical):** Different platforms like Facebook, Instagram, Twitter, TikTok, etc.
3. **Purpose of social media use (Categorical):** Reasons for using social media, such as socializing, news, entertainment, education, etc.
4. **Number of social media accounts (Quantitative):** The total number of different social media accounts a person maintains.
5. **Engagement in social media activities (Quantitative):** The frequency of activities such as posting, liking, commenting, and sharing content.

__Q1.2:__ 

What is the population of the study?\

**Youth ages 18-24**



__Q1.3:__ 

Suppose you are *not* resource limited.  Briefly describe the sampling scheme you would use and why you would choose it to best represent the population:\

**I would use stratified sampling to ensure that different subgroups within the population are adequately represented.**

 
__Q1.4:__ 

Give a brief description of the other 3 types of sampling schemes and explain why they don't make sense in this situation:  

1. **Simple Random Sampling (SRS)**
    * Reason it doesn't make sense: While SRS is straightforward and reduces bias, it may not ensure that all key subgroups within the population are adequately represented.
2. **Systematic Sampling**
    * Reason it doesn't make sense: Systematic sampling might introduce periodicity biases.
3. **Cluster Sampling**
    * Reason it doesn't make sense: Cluster sampling can introduce cluster-specific biases and may not be efficient if clusters (e.g., specific universities or cities) have unique social media usage patterns that aren't representative of the broader population.



__Q1.5:__ 

Suppose it is impractical to sample from the population described in Q1.2.  What population might be more feasible to sample from?\

**A more feasible population to sample from might be college students aged 18-24.**



__Q1.6:__ 

How would modifying the population of interest impact your sampling scheme?\

**Modifying the population of interest would require changing the groups for sampling and adjusting the sample size.**

__Q1.7:__ 

Suppose you sent out a Facebook survey to all of your followers and asked them to respond if they are in your target demographic.  How might this impact the conclusions of your study?\

**Sending a Facebook survey to your followers could introduce self-selection bias which could result in a non-representative sample.**


# Describe/Summarize the Data

For this section you will program in R.  Don't forget to load the libraries first:

```{r, echo=FALSE, include=FALSE}

library(rio)
library(mosaic)
library(tidyverse)
library(car)
library(ggplot2)


```

## Housing Prices in Rexburg

You are interested housing prices in Rexburg, Idaho.  Load the housing data and respond to the following questions.  


```{r}
# Load the data:
housing <- import('https://github.com/byuistats/Math221D_Cannon/raw/master/Data/MadisonCountyRealEstate.xlsx') %>% select(ListPrice, Bedrooms)

```

__Q2.1:__ 

Create a histogram of list price. 

```{r}
ggplot(housing, aes(x = ListPrice)) + 
  geom_histogram(binwidth = 50000, fill = "blue", color = "black") + 
  labs(title = "Histogram of List Prices", x = "List Price", y = "Frequency")
```

What is the shape of the distribution of list prices for homes in Rexburg.\

**The shape of the distribution of list prices for homes in Rexburg is right-skewed, with most prices clustered at the lower end and a few higher-priced homes.**

__Q2.2:__ 

Calculate the mean, standard deviation, 5 number summary and sample size of housing prices for each number of bedrooms.

```{r}
housing_summary <- favstats(ListPrice ~ Bedrooms, data = housing)
knitr::kable(housing_summary)

```

What is the standard deviation of homes with 3 bedrooms?\

**The standard deviation of homes with 3 bedrooms is $79,099.36.**

How do you interpret that number?\

**The standard deviation indicates that the list prices of 3-bedroom homes in Rexburg vary by approximately $79,099.36 around the mean price. This means that the prices of these homes are kind of spread out over a wide range.**


__Q2.3:__

What is the maximum value home price for a 2 bedroom house?\

**$160,000**

Suppose there was a typo in the data and the maximum price for a 2 bedroom house was accidentally entered as $1,700,000.\ 

How would the median change?\

**The median would remain unchanged, as it is not affected by extreme values.**

How would the mean change?\  

**The mean would increase, as it is sensitive to extreme values.**

How would the standard deviation change?\

**The standard deviation would increase, as it measures the spread of the data and would be affected by the large value.**

__Q2.4:__ 

Create a side-by-side boxplot of housing prices for each group.  **Modify the color scheme so that each boxplot is a different color**.  

```{r}
ggplot(housing, aes(x = factor(Bedrooms), y = ListPrice, fill = factor(Bedrooms))) + 
  geom_boxplot() + 
  scale_fill_brewer(palette = "Set3") + 
  labs(title = "Boxplot of List Prices by Number of Bedrooms", x = "Number of Bedrooms", y = "List Price")

```

What trends do you notice?\

**In general, the more bedrooms a house has the more the house costs. The difference in price between 4 and 3 bedroom homes is minimal.**

Why might the box plot for 7 bedroom homes look so different than the others?\

**The box plot for 7-bedroom looks different due to fewer data points and them being outliers.**


# Probability 

## Hair Color

Brother Cannon has been collecting hair color data about his students for the last several semesters.  The proportion of students with each hair color are as follows:

(run the following R chunk to see the table more clearly)

```{r}
knitr::kable(tibble(Red="???", Brown=.37, Black=.35, Blond = .22, Other=.02))
```

__Q3.1:__  

What percent of Brother Cannon's students have red hair?\

**Four percent of students have red hair.**

__Q3.2:__  


What's the probability that a randomly selected student has either black or brown hair?\

**The probability that a randomly selected student has either black or brown hair is 72%.**

## Social Media Use

__Q4.1:__

The number of hours students spend on social media a week is known to be left skewed with a mean of 12 hours and a standard deviation 1.5.  

Suppose we take a random sample of 50 students and calculate the average time they spend on social media.

__Q4.2:__

What is the mean of the distribution of sample means for this sample size?\ 

**The mean is the same as the population mean: 12 hours.**

__Q4.3:__

What is the standard deviation of sample means for this sample size?\  

**The standard deviation of sample means for this sample size would be about .212.**

__Q4.4:__

What is the shape of the distribution of sample means for this sample size and why?\

**The distribution of sample means will be approximately normal due to the Central Limit Theorem.**

## Faculty Salaries

Faculty salaries at a university are known to be normally distributed with a mean of \$108,552 and a standard deviation of \$13,277.

You randomly sample n=15 faculty members and calculate their average salary.  

__Q5.1:__

What is the mean of the distribution of sample means for this sample size?\ 

**The mean is the same as the population mean: $108,552.**

__Q5.2:__

What is the standard deviation of sample means for this sample size?\ 

**The standard deviation of the sample means for a sample size of 15 is approximately $3428.11**

__Q5.3:__

What is the shape of the distribution of sample means for this sample size and why?\ 

**The shape of the distribution of sample means for this sample size is approximately normal because the population from which the samples are drawn is normally distributed.**


## Normal Probability Calculations

High blood pressure is strongly associated with a host of diseases and increases risk of cardiac arrest (hear attacks) in older patients.  

Suppose that systolic blood pressure in the US is normally distributed with a mean = 128 and a standard deviation of 11.  

Use the following calculator to answer the questions below: 

```{r, include=FALSE, echo=FALSE, eval=FALSE}
# Calculator
xbar <- 
mu <- 128
sigma <- 11
n <- 1
sigma_xbar <- sigma / sqrt(n)
sigma_xbar

z <- (xbar - mu) / sigma_xbar
z

# Probability of a sample mean being less than our xbar:
pnorm(z)

# Prob. of a sample mean being higher than our xbar:
1-pnorm(z)

# Prob of being between A and B

A <- 1
B <- 2
  
pnorm(B, mu, sigma_xbar) - pnorm(A, mu, sigma_xbar)

# Percentiles

percentile <- .2
qnorm(percentile, mu, sigma_xbar)


```

__Q6.1:__

What is the probability that a randomly selected person in the US has a systolic blood pressure of greater than 145?\ 

**6.11%**

__Q6.2__

What is the probability that a randomly selected group of 10 people in the US have an *average* blood pressure less than 130?\ 

**71.73%**

__Q6.3:__

Your doctor says you're in great shape, and that you are in the 20th percentile for systolic blood pressure.

What does it mean that you are the 20th percentile?\ 

**Being in the 20th percentile means that your systolic blood pressure is lower than 80% of the population.**

What is your blood pressure?\

**My blood pressure would be approximately 118.74.**

