---
title: "Paired T-Test"
subtitle: "Homework Quiz"
author: "Nathan Lunceford"
warning: false
format:
  html:
    self-contained: true
    code-fold: true
---

# Helmet Measurements

After purchasing a batch of flight helmets that did not fit the heads of many pilots, the NZ Airforce decided to measure the head sizes of all recruits. Before this was carried out, information was collected to determine the feasibility of using cheap cardboard calipers to make the measurements, instead of metal ones which were expensive and uncomfortable. 

The data lists the head diameters of 18 recruits measured once using cardboard calipers and again using metal calipers. One question is whether there is *any* systematic difference between the two measurement tools.

__NOTE__:  Define the difference as cardboard - metal. 

Use $\alpha = 0.05$ significance level.

## Step 1: Load the Data


```{r}
library(tidyverse)
library(mosaic)
library(rio)
library(car)

helmet_fit <- import("https://byuistats.github.io/M221R/Data/quiz/R/helmet_fit.csv")

```

## Step 2: Explore the Data and Generate Hypotheses

Create histograms and summary statistic tables for the measurements for each type of tool:\
__Cardboard__
```{r}
histogram(helmet_fit$cardboard)
fav <- favstats(helmet_fit$cardboard)
fav

```
__Metal__
```{r}
histogram(helmet_fit$metal)
fav <- favstats(helmet_fit$metal)
fav

```

## Step 3: Prepare the data for analysis

Give the summary statistics (favstats) for the *differences* in the measured head diameters.  

What does a negative number mean given the definition of the difference?

```{r}
diff <- helmet_fit$cardboard - helmet_fit$metal
fav <- favstats(diff)
fav
```
__A negative number in the differences means that the metal caliper measurement is larger than the cardboard caliper measurement.__

Create a histogram and a qqPlot of the differences:

```{r}
histogram(diff)
qqPlot(diff)

```



## Step 4: Perform the Appropriate Analyses

### Hypothesis Test

Make the following null and alternative hypothesis correct by deleting what doesn't belong: 

$$H_0: \mu_d  = 0$$

$$ H_a: \mu_d \ne 0$$

```{r}
# Perform a t-test for the mean of the differences between cardboard and caliper data
t_test_results <- t.test(diff, mu = 0, alternative = "two.sided")

# Print the test statistic, degrees of freedom, and p-value
t_test_results


```

What is the value of the test statistic, $t$?\
__t = 3.1854__

How many degrees of freedom does this test statistics have?\
__df = 17__

What is the P-value?\
__p-value = 0.005415__

Can we trust this P-value? 

  a. How many differences in our sample?\
  __18__
  b. Check the `qqPlot()` of the differences for normality:
  __We can assume normality because the qqPlot shows that the points generally follow a straight line.__


State your conclusion about the hypothesis test:\
__Since the p-value (0.005415) is less than the significance level, we reject the null hypothesis. This indicates that there is a statistically significant difference between the head diameter measurements taken with cardboard calipers and metal calipers.__



### Confidence Intervals

Create a 95% confidence interval for the true average difference between the cardboard and the metal measurement tools:

```{r}
conf_int <- t_test_results$conf.int
conf_int
```


Give a one-sentence explanation of your confidence interval:\
__We are 95% confident that the true mean difference in head diameter measurements between the cardboard and metal calipers lies between 0.544 and 2.678.__



# Cholesterol

A study was conducted at a major northeastern American medical center regarding blood cholesterol levels and heart-attack incidents. A total of 28 heart-attack patients had their cholesterol levels measured two days after the heart attack and again four days after the heart attack.  Researchers want to see if cholesterol levels reduced significantly four days after the attack versus two days after the attack.  

Use the level of significance of 0.05.


## Step 1:  Load the Data

```{r}

cholesterol <- import("https://byuistats.github.io/M221R/Data/quiz/R/cholesterol.csv")
glimpse(cholesterol)
```

## Step 2:  Review the Data Generate Hypotheses

Create histograms and summary statistics tables for the cholesterol measurements at 2 days and at 4 days.
__Cholesterol 2 Days__
```{r}
histogram(cholesterol$chol_day2)
fav <- favstats(cholesterol$chol_day2)
fav

```
__Cholesterol 4 Days__
```{r}
histogram(cholesterol$chol_day4)
fav <- favstats(cholesterol$chol_day4)
fav

```

## Step 3:  Prepare the data for analysis

Decide how you are going to define the difference (chol_day2 - chol_day4 or chol_day4 - chol_day2).\
__I will define the difference as chol_day2 - chol_day4.__

What does a negative number mean:\
__A negative number means that the cholesterol level is lower at 4 days compared to 2 days, indicating a reduction.__

What is your null and alternative hypotheses:\


$$H_0: \mu_d  = 0$$


$$ H_a: \mu_d < 0$$

Create a qqPlot of the differences and determine if you can trust the statistical tests.

```{r}
diff <- cholesterol$chol_day2 - cholesterol$chol_day4
qqPlot(diff)
```

## Step 4:  Perform the appropriate analysis


### Hypothesis Test

Perform a t-test for the differences.

```{r}
t_test_results <- t.test(diff, mu = 0, alternative = "less")

t_test_results
```

How many degrees of freedom does the test statistic have:\
__df = 27__

What is the test statistic:\
__t = 3.2189__

What is the P-value:\
__p-value = 0.9983__

State your conclusion in context of the research question:\
__Given the p-value (0.9983), which is greater than alpha, we fail to reject the null hypothesis. This means that there is not enough evidence to suggest that cholesterol levels are significantly lower four days after the heart attack compared to two days after the heart attack.__


### Confidence Interval

Create a 95% confidence interval for the difference in cholesterol scores:

```{r}
conf_int <- t_test_results$conf.int
conf_int

```


Explain your confidence interval in context of the research question:\
__The 95% confidence interval for the difference in cholesterol levels between 2 days and 4 days is (−∞,35.60725) suggesting that we cannot conclude a significant reduction in cholesterol levels from day 2 to day 4.__




