---
title: "Introducing the Bivariate Data"
subtitle: "A Better Way to Visualize Data"
format:
  html:
    self-contained: true
    code-fold: true
warning: false
---

```{r}

library(tidyverse)
library(mosaic)
library(rio)

geyser <- import('https://byuistats.github.io/BYUI_M221_Book/Data/OldFaithful.xlsx')
#names(geyser)

math <- import('https://byuistats.github.io/BYUI_M221_Book/Data/MathSelfEfficacy.xlsx')
#names(math)


real_estate <- import('https://github.com/byuistats/Math221D_Cannon/raw/master/Data/MadisonCountyRealEstate.xlsx')
body_measures <- import('https://github.com/byuistats/Math221D_Cannon/raw/master/Data/BodyMeasurementsCorrected.xlsx')
```


# Correlation Coefficients

What does the correlation coefficient, r, quantify?

What does it mean if the correlation coefficient, r, is negative?


# Geyser Data

## Scatter plot

Make a scatter plot showing the relationship between wait time and the duration of the next eruption.

Which variable is the Explanatory variable?  Which is the Response?

```{r}

#plot()

```

Before looking at the Correlation Coefficient, r, describe in words the direction and strength of the relationship.  Does it look linear?

## Correlation Coefficient

```{r}

#cor()

```



# Geyser Data

## Scatter plot

Make a scatter plot showing the relationship between the self reported confidence rating and their test score.

Which variable is the Explanatory variable?  Which is the Response?


```{r}

#plot()

```

Before looking at the Correlation Coefficient, r, describe in words the direction and strength of the relationship.  Does it look linear?

## Correlation Coefficient

```{r}

#cor()

```


```{r}
plot(real_estate$Age, real_estate$ListPrice,
     main = "Scatterplot of Home Prices vs. Age of Homes",
     xlab = "Age of Homes",
     ylab = "List Price of Homes",
     pch = 19,  # Plotting character (solid circle)
     col = "blue")  # Color of the points
```



```{r}
correlation <- cor(body_measures$Height, body_measures$Weight)
correlation
```


```{r}
plot(body_measures$Height, body_measures$Weight,
     main = "Scatterplot of Height vs. Weight",
     xlab = "Height (inches)",
     ylab = "Weight (pounds)",
     pch = 19,  # Plotting character (solid circle)
     col = "blue")  # Color of the points
```

```{r}
# Given values
mu <- 1.59
sigma <- 0.042
n <- 20
x_bar <- 1.57

# Compute the standard error of the mean
standard_error <- sigma / sqrt(n)

# Calculate the Z-score
z_score <- (x_bar - mu) / standard_error

# Calculate the probability
probability <- pnorm(z_score)

# Print the result
print(probability)
```

