---
title: "Summarizing Quantitative Data"
subtitle: "Homework"
format:
  html:
    self-contained: true
    code-fold: true
warning: false
---


# Load the libraries
```{r}

library(rio)
library(mosaic)
library(tidyverse)
library(car)

```

# Read in Data

```{r}

old_faithful <- rio::import("https://byuistats.github.io/M221R/Data/old_faithful.xlsx")
view(old_faithful)
gratitude <- rio::import("https://byuistats.github.io/M221R/Data/gratitude.xlsx")
view(gratitude)

ext_fav <- favstats(old_faithful$duration, na.rm = TRUE)
```


# Old Faithful

## Calculate the Summary Statistics for Duration

```{r}
duration_favstat <- favstats(old_faithful$duration, na.rm = TRUE)

```

What is the mean duration time of Old Faithful eruptions?\
3.488 units\
What is the standard deviation of duration?\
1.141 units


## Create a Historgram for Duration

Create a histogram and describe the shape of the distribution of duration:

```{r}
histogram(old_faithful$duration)


```

The shape is bi-modal.

## Calculate Summary Statistics for Wait time

```{r}
wait_favstat <- favstats(old_faithful$wait, na.rm = TRUE)
wait_favstat

```

What is the mean wait time between eruptions?\
70.897 units\
What is the maximum wait time between eruptions?\
96 units\
The middle 50% of wait times will be between what 2 numbers?\
The middle 50% will be between Q1 and Q3.


# Gratitude

Calculate the summary statistics for both groups:

```{r}
fav <- favstats(gratitude$happiness ~ gratitude$treatment)
fav
#interested category then column you want to break it up by.
```

Which treatment group has the highest mean?\
Grateful\
Which treatment group has the largest spread of happiness scores?\
Grateful\
Create a graph with boxplots for each group:

```{r}
boxplot(gratitude$happiness ~ gratitude$treatment)

```

Which group has the lowest first quartile?\
Hassles


